\documentclass[12pt]{article}

\usepackage{a4wide, amsfonts, epsfig}

\begin{document}
\begin{center}
{\bf EMAT10001 Exercise Sheet 16.}\\[1cm]{} Conor Houghton 2014-02-18
\end{center}
\subsubsection*{Introduction} 
This worksheet is about multidimensional calculus and the gradient
vector. There is the usual bounty for errors and typos, 20p to \pounds
2 depending on how serious it is.

\subsubsection*{Exercise sheet}

The difference between the work sheet and the exercise sheet is that
the solutions to the exercise sheet won't be given and the problems
are designed to be more suited to working on on your own, though you
are free to discuss them in the work shop if you finish the work sheet
problems. Selected problems from the exercise sheet will be requested
as part of the continual assessment portfolio.

\begin{enumerate}


\item Find $\partial f/\partial x$ and $\partial f/\partial y$ for 
\begin{enumerate}
\item $f(x,y)=xy\ln{xy}$
\item $f(x,y)=12x^4y+y^2$
\item $f(x,y)=xe^{y^2}$
\end{enumerate}


\item Find $\partial f/\partial x$, $\partial f/\partial y$ and $\partial f/\partial z$ for 
\begin{enumerate}
\item $f(x,y,z)=xyz$
\item $f(x,y,z)=(x-y)^2+(y-z)^2+(z-x)^2$
\end{enumerate}

\item For $f(x,y)=xy$ work out the directional
  derivative in the $(1,3)$ direction at $(1,1)$; don't forget to
  normalize the direction vector.

\item The third differential operator is curl; it acts on vector
  fields to give another vector field. It is only defined in three
  dimensions and has quite a complicated form
\begin{equation}
\mbox{curl}\mathbf{v}=\left(\frac{\partial v_3}{\partial y}-\frac{\partial v_2}{\partial z},\frac{\partial v_1}{\partial z}-\frac{\partial v_3}{\partial x},\frac{\partial v_2}{\partial x}-\frac{\partial v_1}{\partial y}\right)
\end{equation}
Show grad$\,(\mbox{curl}\mathbf{v})=0$.

\end{enumerate}

\subsubsection*{Challenge}
First four to email or tell me the answer:
\begin{eqnarray}
&&>+++++[<++>-]<+++ >++++[>+++<-]>[<++++>-]\cr 
&&<.+< . >. < . >-> >\cr
&&+++++>+< [ >[> >+>+< < <-] > > >[< < <+> > >-]< < < >[<+>-]\cr
&&< < < <[>+> > > > >+< < < < < <-] >[<+>-] > >\cr
&&[< <+> > > > >+< < <-] < <[> >+< <-]> > > > >.[-]\cr 
&&< < < < < < < .  > > > > > > [<+>-]< < <-]
\end{eqnarray}
The start of what?
 \end{document}
